Using species distribution models to direct native seed collection efforts

### 2024-09-11 scripts/Manuscript.Rmd
abstract

--- 
title: 'Using Species Distribution Models to inform native seed collection efforts' 
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  less than 250 words - no subsections.           
keywords: |
  species distribution models, native seed development  
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/restoration-ecology.csl"
bibliography: ../citations/bibliography/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating} 
---

##  IMPLICATIONS FOR PRACTICE

up to six of these. let's shoot 3 with 2 sentence max per.

@@

With the United Nations 'Decade of Restoration' well underway limitations in our capacity to carry out active restoration, which normally requires large volumes of seed, are becoming apparent [@national2023assessment]. 
Given the scale of our restoration needs, in most scenarios, the only sustainable source of this quantity of seed are from the amplification ('grow-out') of wild harvested seed in agricultural settings [@pedrini2020collection; @broadhurst2015seeding; @national2023assessment].
Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available from agricultural amplification for restoration projects [@national2023assessment, @merritt2011restoration]. 
However, numerous difficulties exist in both the wild harvest of seed and it's subsequent  amplification which are limiting our ability to develop adequate amounts of germplasm.

@@

A Species Distribution Model (alternatively referred to as 'Environmental niche models' etc.) uses presence, and optionally absence, records of a species and correlates these records to environmental variables (e.g. climate, land cover, and soil), using various families of statistical models (e.g. GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface. 
While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@yee). 
In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g. they proliferate on uncommon soils and are outcompeted on more typical soil types. 
Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored. 
Testing the applicability of SDMs to modeling the range of common species is further valuable as much biodiversity planning incorporates common species which are oftentimes viewed as the fabric holding together ecosystems, and as they may assist with planning restoration plantings under climate change.

@@

## Species Occurrence Records  
Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio]. 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (bio1, 4, 10, 12, & 19; see TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

Species absences were generated using three processes, and we sought to have a 1:1 ratio of presences to absences. 
True absences were acquired from an ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g. if 20% of land ownership across a species range was BLM, than the number of presences * 0.2 defined the number of these absences). 
Likely absences, representing 15% of records, were generated outside the known range of the species, but bounded to be within 50 miles of the extent extent of occurrences. 
Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for ((1 - (% BLM Land + 15 % LA records)) * 1.25).
Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used. The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables [@venables2002mass], identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables [@naimi2014usdm]. 
As many records, classified by the LDA as originating from the Presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.

## Environmental Variables  
Up to 44 variables were used in generating the species distribution models (Appendix 1). 
These variables were selected based on authors previous work, and represent variables commonly associated with the empirical distributions of taxa in arid and semi-arid regions. 
The thematic contents of these variables largely relate to climate [@karger2017climatologies], landcover [@tuanmu2014global], topography and landform [@amatulli2020geomorpho90m ; @yamazaki2017merit], soil [@hengl2017soilgrids250m; @ivushkin2019global], and representations of anthropogenic impacts [@sanderson2002human]. 
All variables were downloaded from source and re-sampled to 90m resolution using terra [@hijman2024terra].

## Species Distribution Modelling  
Random Forests models were generated for each species following several steps to reduce the number of features.
While superfluous features generally do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables [@kursa2010boruta], and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.  
Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys [@kuhn2008caret; @liaw2002randomForest]. 
Models were then predicted onto raster surfaces which exceeded the species known range by 50 miles [@hijman2024terra].

## Patch identification 
To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more [@usgs2004nhd], and the divides of HU10 watersheds [@usgs2023wbd] were 'burnt' away from the raster. 
The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a practical time period. 
Resultant patches < 5 acres were then discarded, and the rasters was resampled to it's input resolution.

>>>>

--- 
title: 'Using species distribution models to direct native seed collection efforts' 
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  Developing many locally adapted germplasms for habitat restoration has been a challenging process. 
  In particular, finding a large enough number of populations which can serve as source material for agricultural increase across the range of species is proving difficult. 
  The population scouting efforts of native seed collection crews can be informed using Species Distribution Models, allowing them to prioritize areas with higher probabilities of successfully finding the target species. 
  However, SDM's themselves do not have a mechanism for addressing the discontuity between suitable habitat and dispersal limitation, requiring that some of connectivity analysis be performed as a post-processing step to aid in their interpretation. 
  Herein we detail the high-throughput and semi-automated development of SDM's for over 350 common taxa native to Western North America. 
  We further employ simple connectivity analysis, based solely on drainage basins and neighborhood analysis, on these hypothesis to try and ensure that areas with higher probabilities of propagule dispersal and population establishment are visited. 
  Using the hypothesis of suitable habitat generated by these hypothesis, with ground verification data from 13 crews in 5 Western states, we show that SDMs are useful for crews to find populations of native species, and that they are most useful when combined with connectivity analysis. 
  Our workflow is readily replicable by users on a variety of low-cost computing systems, and our implementation strategies for field crews are briefly discussed. 
keywords: |
  species distribution models, native seed development  
output:
  pdf_document: default 
  toc: no 
  word_document: default 
csl: "../citations/restoration-ecology.csl" 
bibliography: ../citations/bibliography/citations.bib 
link-citations: yes 
fig_caption: yes
always_allow_html: yes 
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating} 
---

##  Implications for Practice

- Finding enough large populations across seed collection zones (e.g. seed transfer zones) for economically viable agricultural increase has proven to be difficult. 
- Heuristic tools which can be generated cheaply and computationally such as Species Distribution Models, and connectivity analysis, offer a possibility to increase the number of populations eligible for wild land seed collection. 
- Our code and protocol for generating and utilizing these tools are available and widely applicable to other users.

@@

With the United Nations 'Decade of Restoration' well underway limitations in our capacity to carry out active restoration, which normally requires large volumes of seed, are becoming apparent [@national2023assessment]. 
Given the scale of our seed-based restoration needs, in most scenarios, the only sustainable source of this quantity of seed are from the amplification ('grow-out') of wild harvested seed in agricultural settings [@pedrini2020collection; @broadhurst2015seeding; @national2023assessment].
Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available from agricultural amplification for restoration projects [@national2023assessment, @merritt2011restoration]. 
However, numerous difficulties exist in both the wild harvest of seed and it's subsequent  amplification which are limiting our ability to develop adequate amounts of germplasm.

@@

A Species Distribution Model (alternatively referred to as 'Environmental niche models' etc.) uses presence, and optionally absence, records of a species and correlates these records to environmental variables (e.g. climate, land cover, and soil), using various families of statistical models (e.g. GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface. 
While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@a2022species). 
In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g. they proliferate on uncommon soils and are outcompeted on more typical soil types. 
Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored. 
Testing the applicability of SDMs to modeling the range of common species is further valuable as much biodiversity planning incorporates common species which are oftentimes viewed as the fabric holding together ecosystems, and as they may assist with planning restoration plantings under climate change.

@@

## Species Occurrence Records  
Species presence records were collected from herbaria, citizen science initiatives (e.g. iNaturalist), and standardized ecological monitoring programmes (e.g. VegBank, Forest Inventory and Analysis) using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio]. 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

Species absences were generated using three processes, and we sought to have a 1:1 ratio of presences to absences. 
True absences were acquired from an ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g. if 20% of land ownership across a species range was BLM, than the number of presences * 0.2 defined the number of these absences). 
Likely absences, representing 15% of records, were generated outside the known range of the species, but bounded to be within 50 miles of the extent extent of occurrences. 
Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for $((1 - (\text{\% BLM Land} + 15 \text{\% LA records})) * 1.25)$.
Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used. The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables [@venables2002mass], identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables [@naimi2014usdm]. 
As many records, classified by the LDA as originating from the presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.

## Independent Variable Processing 
Up to 44 variables were used in generating the species distribution models (Appendix 1). 
These variables were selected based on authors previous work, and represent variables commonly associated with the empirical distributions of taxa in arid and semi-arid regions. 
The thematic contents of these variables largely relate to climate [@karger2017climatologies], landcover [@tuanmu2014global], topography and landform [@amatulli2020geomorpho90m ; @yamazaki2017merit], soil [@hengl2017soilgrids250m; @ivushkin2019global], and representations of anthropogenic impacts [@sanderson2002human]. 
All variables were downloaded from source and re-sampled to 90m resolution using terra [@hijman2024terra].

## Species Distribution Modelling  
Random Forests models were generated for each species following several steps to reduce the number of features.
While superfluous features do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
In our experience, predicting models onto raster surfaces is the most time consuming step of SDM generation. 
The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables [@kursa2010boruta], and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33 [@naimi2014usdm].  
Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling, using the 'randomForest' package with all default values except for optimized mtrys [@kuhn2008caret; @liaw2002randomForest]. 
Models were then predicted onto raster surfaces which exceeded the species known range, based on the training and test data, in all directions by 50 miles [@hijman2024terra].

## Patch identification 
To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more [@usgs2004nhd], and the divides of HU10 watersheds [@usgs2023wbd] were 'burnt' away from the raster. 
The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a scalable time period. 
Resultant patches < 5 acres were then discarded, and the rasters was resampled to it's input resolution.
